### 2008-07-03 trunk/Papers/rmrs-08/abstract.tex
3rd July

\begin{abstract}
  la la la
\end{abstract}

>>>>

\begin{abstract}
  One way to construct semantic
  representations in a robust manner is to enhance shallow language
  processors with semantic components (e.g., \cite{frank:2004,
    briscoe:etal:2006}).  Here, we provide a model theory for a
  semantic formalism 
  that is designed to enable this,
  namely Robust Minimal Recursion Semantics ({\sc rmrs},
  \cite{copestake:2003}).  We show that {\sc rmrs} supports a notion
  of entailment 
  that has a number of desirable properties: specifically, it allows
  {\sc rmrs} to form the basis for comparing the semantic output of
  different parses of the same string, even when the parses vary in
  the depth of their analysis, and in particular to test the
  compatibility of the semantic output of a shallow parser with that
  of a deep parser (where such deep parses exist).
\end{abstract}

### 2008-07-03 trunk/Papers/rmrs-08/introduction.tex
3rd July

Representing semantics as a logical form that supports automated
inference and model generation is vital for deeper language
engineering tasks, such as dialogue systems.  Such detailed semantic
representations can be obtained from hand-crafted deep grammars (e.g.,
\cite{butt:etal:1999,copestake:flickinger:2000}), but deep parsing
tends to lack robustness: hand-crafted grammars fail to cover all
words and linguistic constructions; and unedited text and speech
contains ill-formed phrases, which by design deep grammars do not
handle.

On the other hand, there has been a recent trend towards systems for
robust wide-coverage semantic construction.  For instance,
\cite{bos:etal:2004} build upon a statistical CCG parser, exploiting
the elegant syntax-semantics interface that CCG provides;
\cite{zettlemoyer:collins:2007} and others combine syntactic analyses
with shallow features to learn string-to-semantics mappings; and
\todo{say something about RASP and Anette's work}.

However, there are certain semantic phenomena that such shallow
approaches can't capture, including quantifier scope, optional
arguments, and long-distance dependencies.  Forcing a shallow parser
to make a decision about these phenomena is therefore a major source
of errors.  It would be preferrable to give the parser the option to
declare itself not sufficiently informed to make the decision, and
leave the decision open -- i.e., to compute a partial semantic
representation that neither over-determines nor under-determines the
semantic information revealed by the (shallow) syntactic analysis, and
can later be completed by further inferences.

In this paper, we focus on one particular approach to representing
partial semantic information and producing it from robust parsers:
Robust Minimal Recursion Semantics ({\sc rmrs},
\cite{copestake:2003}).  {\sc rmrs} is a generalisation of {\sc mrs}
\cite{copestake:etal:2005} designed to support underspecification both
of scope relations and of the predicate-argument structure, and has
been demonstrated to support semantic constructions of shallow parsers
ranging in depth from {\sc pos} taggers (refs) to chunk parsers (frank
ref) and intermediate parsers ({\sc rasp} ref).  {\sc mrs} analyses
derived from deep grammars, such as the English Resource Grammar
\cite{copestake:flickinger:2000} can also be seen as special cases of
{\sc rmrs} representations.

>>>>

Representing semantics as a logical form that supports automated
inference and model construction is vital for deeper language
engineering tasks, such as dialogue systems.  Such detailed semantic
representations can be obtained from hand-crafted deep grammars (e.g.,
\cite{butt:etal:1999,copestake:flickinger:2000}), but deep parsing
tends to lack robustness: hand-crafted grammars fail to cover all
words and linguistic constructions; and unedited text and speech
contains ill-formed phrases, which by design deep grammars do not
handle.

On the other hand, there has been a recent trend towards systems for
robust wide-coverage semantic construction.  For instance,
\cite{bos:etal:2004} exploit
{\sc ccg}'s elegant syntax-semantics interface to produce a unique
logical form for each derivation of
a statistical {\sc ccg} parser;
\cite{zettlemoyer:collins:2007} and others learn
string-to-logical-form mappings, exploting informative features from
the output of
shallow language processors; and \cite{copestake:2003} and
\cite{frank:2004}, among others,
endow robust language processors with semantic components.

However, there are certain semantic
phenomena that such shallow
approaches can't capture reliably, including quantifier scope, optional
arguments,
% COMMENT: this issue about optional arguments is really too cryptic.
% Basically, there is no way for assessing by *any* learned language
% processor whether a single word form that appears in
% a number of different syntactic valencies is an example of lexical
% sense ambiguity, or they are all examples of the same sense being
% used in syntactically different ways (i.e., we have optional
% arguments).  Only hand-crafted grammars can tell you this.  So, for
% example, Bos et al (2004) would get bet(x,y) for "I bet 5 pounds",
% and bet(x,y,z) for "I bet you 5 pounds" and bet(x,p) for "I bet it's
% going to rain": I.e., the vocab of the semantic formalism is
% constructed automatically from NL word forms, and the predicate
% symbols in that automatically constructed vocab vary in (a) valency,
% and (b) the semantic dependencies represented by a given argument
% position (e.g., the second argument of "bet" in "I bet 5 pounds" is
% 5 pounds, and for "I bet it's raining" it's "it's raining".
% 
and long-distance dependencies.  
% TODO: search Steedman and Clark EMNLP 2004 for the performance
% figures on object relative clauses (which is constructions like
% "The man that I met", "the man that I think likes me", but not "the
% man that likes me"; they get about 50% of them exactly right; this
% is impressive but shows it's error prone.
Forcing a shallow parser
to make a decision about these phenomena therefore can be a major source
of errors.  It would be preferrable to give the parser the option to
declare itself not sufficiently informed to make the decision, and
to leave the decision open -- i.e., to compute a partial semantic
representation that neither over-determines nor under-determines the
semantic information revealed by the (shallow) syntactic analysis, and
to complete the semantic information later, via inferences that
exploit information extraneous to the shallow parser.

In this paper, we focus on one particular approach to representing
partial semantic information:
Robust Minimal Recursion Semantics ({\sc rmrs},
\cite{copestake:2003}).  {\sc rmrs} is a generalisation of {\sc mrs}
\cite{copestake:etal:2005}, designed to support underspecification both
of scope relations and of the predicate-argument structure.
\cite{copestake:2003, frank:2004} use it to
support semantic constructions by shallow parsers
ranging in depth from {\sc pos} taggers to chunk parsers  
and intermediate parsers such as {\sc rasp}
\cite{briscoe:etal:2006}.  {\sc mrs} analyses
derived from deep grammars, such as the English Resource Grammar
\cite{copestake:flickinger:2000} can also be seen as special cases of
{\sc rmrs} representations.

### 2008-07-03 trunk/Papers/rmrs-08/motivation.tex
3rd July

We motivate and describe the main features of {\sc rmrs} with a (toy)
example.  Sentence 
(\ref{eg:cat}a) features a semantic scope ambiguity that, according to
most deep grammars, syntax
doesn't resolve.    The {\sc mrs} (\ref{eg:cat}b) derived by the
English Resource Grammar ({\sc erg}, \cite{copestake:flickinger:2000})
reflects this by underspecifying socpe: (i) each predication is
labelled ($l_1$, $l_4$, etc); (ii) scopal arguments are indicated by
holes (e.g., $h_2$ and $h_3$ in the quantifier {\em every}); and (iii)
constraints on scope are expressed by the qeq conditions (roughly, $h
=_q l$ means that only zero or more quantifiers lie between the scopal
argument $h$ and the predication labelled by $l$ in any fully-resolved
logical form); and (iv) these scope
constraints allow for more than one way of `plugging' the holes with
labels.\footnote{The event variable in the predicate symbol
  corresponding to the adjective is motivated by copulas---{\em The
    cat is fat} arguably has the logical form 
  $\mbox{\_the\_q\_1}(x,\mbox{\_cat\_n\_1}(x),\mbox{\_fat\_j\_1}(e,x))$.}
\begin{examples}
\item   \label{eg:cat}
\begin{subexamples}
\item
Every fat cat chased some dog.
% \item
% $\mbox{\_every\_q\_1}(x, \mbox{\_fat\_j\_1}(e',x) \wedge
% \mbox{\_cat\_n\_1}(x),$\\
% \hspace*{0.5in}$\mbox{\_some\_q\_1}(y, \mbox{\_dog\_n\_1}(y),
% \mbox{\_chase\_v\_1}(e_{spast},x) )$\\
% $\mbox{\_some\_q\_1}(y, \mbox{\_dog\_n\_1}(y),$\\
% \hspace*{0.5in}$\mbox{\_every\_q\_1}(x, \mbox{\_fat\_j\_1}(e',x)
% \wedge 
% \mbox{\_cat\_n\_1}(x), 
% \mbox{\_chase\_v\_1}(e_{spast},x)))$
\item   $l_1\handel\mbox{\_every\_q\_1}(x,h_2,h_3)$\\
$l_4\handel\mbox{\_fat\_j\_1}(e',x)$\\
$l_4\handel\mbox{\_cat\_n\_1}(x)$\\
$l_5\handel\mbox{\_some\_n\_1}(y,h_6,h_7)$\\
$l_8\handel\mbox{\_dog\_n\_1}(y)$\\
$l_9\handel\mbox{\_chased\_n\_1}(e,x,y)$\\
$h_0 =_q l_9, h_2 =_q l_4, h_6 =_q l_9$
\end{subexamples}
\end{examples}
(\ref{eg:cat}b) also exhibits {\sc mrs}' compact description of
conjunction through label sharing (see $l_4$ above), and the {\sc
  erg}'s naming convention for predicate symbols: each predicate is
constructed from the word lemma, its syntactic category and a sense
number, and the leading underscore identifies lexical predicates.

Conceptually, this {\sc mrs} is a partial description of all its
possible (fully specific) interpretations in context.  
These are expressed in a so-called {\em base language}, where each
predicate symbol in the {\sc mrs} is a proxy for one or more
base-language constructors of the same arity
Assuming a very close
correspondence between the {\sc mrs} predicate symbols and the
base-language constructors they 
stand for, the {\sc mrs}
(\ref{eg:cat}b) (partially) describes both the base-language formulae in
Figure~\ref{fig:1}, shown as trees to emphasise that an {\sc mrs} is a
partial description of the formulae's {\em form}; not its
(base-language) entailments. 
In essence, the labels in the {\sc mrs} reify the
nodes in the tree, and each of these trees encodes one way of equating
holes with labels that respects qeq constraints and binding conditions
on the variables.

@@

The {\sc mrs} (\ref{eg:cat}c) is a notational variant of the {\sc
  rmrs} (\ref{eg:cat}d). {\sc rmrs}s offer a more factorised
representation: its predicate symbols are all unary and the other
arguments to the (base-language) constructor that it corresponds to
are represented by separate binary relations on the unique {\em
  anchor} of the predicate symbol ($a_1, a_{41},\ldots$) together with
variable and label equalities (e.g., $x=x_1$, $l_{41}=l_{42}$).
\begin{examples}
\item   [\ref{eg:cat}]
\begin{subexamples}
\item   [c]
$l_1:a_1\handel\mbox{\_every\_q\_1}(x_1),$\\
\hspace*{0.1in}$\mbox{RSTR}(a_1,h_2),
\mbox{BODY}(a_1,h_3)$\\ 
$l_{41}:a_{41}\handel\mbox{\_fat\_j\_1}(e'), \mbox{ARG1}(a_{41},x_2)$\\
$l_{42}:a_{42}\handel\mbox{\_cat\_n\_1}(x_3)$\\
$l_5:a_5\handel\mbox{\_some\_q\_1}(x_4)$,\\
\hspace*{0.1in}$\mbox{RSTR}(a_5,h_6),
\mbox{BODY}(a_5,h_7)$\\ 
$l_8:a_8\handel\mbox{\_dog\_n\_1}(x_5)$\\
$l_9:a_9\handel\mbox{\_chase\_v\_1}(e_{spast})$,\\
\hspace*{0.1in}$\mbox{ARG1}(a_9,x_6),
\mbox{ARG2}(a_9,x_7)$\\ 
$h_2=_q l_{42}, l_{41}=l_{42}, h_6 =_q l_8, h_0=_q l_9$\\
$x_1=x_2,x_2=x_3, x_4=x_5,$\\
$x_3=x_6, x_5=x_7$
\end{subexamples}
\end{examples}

This factored representation allows one to build semantic components
to shallow parsers where lexical or syntactic information is absent.
An extreme example would be a {\sc 
  pos} tagger: one can build its semantic component simply by deriving
lexical predicate symbols from the word lemmas and their tags and ensuring
that the predication for each word receives a unique label, anchor and
argument, as given in (\ref{eg:cat-pos}):
\begin{examples}
\item   \label{eg:cat-pos}
\begin{subexamples}
\item   Every\_{\sc at1} fat\_{\sc jj} cat\_{\sc nn1} chased\_{\sc
    vvd} some\_{\sc dd} dog\_{\sc nn1}
\item   $l_1:a_1:\mbox{\_every\_q}(x)$, \\
$l_{41}:a_{41}:\mbox{\_fat\_a}(e')$,\\
$l_{42}:a_{42}:\mbox{\_cat\_n}(x_3)$\\
$l_9:a_9:\mbox{\_chased\_v}(e_{past})$, \\
$l_5:a_5:\mbox{\_some\_q}(x_4)$, \\
$l_8:a_8:\mbox{\_dog\_n}(x_5)$
\end{subexamples}
\end{examples}
Semantic relations, sense tags and the arity of the base-language
constructors are
missing from (\ref{eg:cat-pos}b) because the {\sc pos} tagger doesn't
reveal information about syntactic constituency, word sense or lexical
subcategorisation respectively.  
But the {\sc rmrs}s (\ref{eg:cat}c) and
(\ref{eg:cat-pos}b) are entirely compatible, the former being more
specific than the latter.  At least, this holds so long as the
base-language constructors that the {\sc rmrs} predicate symbol
{\_name\_t\_s} corresponds to is a 
subset of those corresponding to the predicate symbol {\em \_name\_t},
for any word lemma {\em name} and {\sc pos} tag $t$ (written 
$\mbox{{\em \_name\_t\_s}} \sqsubset \mbox{{\em name\_t}}$).

The values of non-scopal arguments are underspecified through the
absence of ARG relations and/or through the absence of
equalities and inequalities among the variables $x_1, x_2,\ldots$.
The need to
underspecify such semantic dependencies is required when
the shallow parser doesn't record syntactic dependencies.
Indeed, there is no single {\sc mrs} which is satisfied by both the
trees in Figure~\ref{fig:2}, assuming of course that disjunction is not part of
the {\sc mrs} language.  In words, these trees collectively specify
the (partial) semantic information that $x$ is either the second
argument (i.e., the direct object) or the third argument (i.e., the
indirect object) to {\em give}, but we don't know which.  
A shallow
processor that misses long distance dependencies in
(\ref{eg:which-dog}) cannot
discern that the bone is the second argument to {\em
  give} in (\ref{eg:which-dog}a), while the dog is the third argument
in (\ref{eg:which-dog}b).
\begin{examples}
\item   \label{eg:which-dog}
\begin{subexamples}
\item   To which dog did Kim give the bone?
\item   Which bone did Kim give the dog?
\end{subexamples}
\end{examples}
Thus to maximise the semantic contribution
of the parser while not overdetermining it, one should express that
the argument position of the noun phrases in {\em give the bone} vs.\
{\em give the dog} are underspecified between the second and third
position, as shown in Figure~\ref{fig:2}.   This will be expressed in
{\sc rmrs} via the binary relation  $\mbox{{\em ARG}}_{\{2,3\}}(a,x)$
between the anchor $a$ of the predicate symbol for {\em give} and the
individual $x$.   More generally, $\mbox{{\em ARG}}_n(a,x)$ means that
$x$ is an argument to the predicate anchored at $a$, but its argument
position is unspecified.

>>>>

We motivate and describe the main features of {\sc rmrs} with a toy
example.  Sentence 
(\ref{eg:cat}a) features a semantic scope ambiguity that, according to
most deep grammars, syntax
doesn't resolve.    The {\sc mrs} (\ref{eg:cat}b) derived by the
English Resource Grammar ({\sc erg}, \cite{copestake:flickinger:2000})
reflects this: (i) each predication is
labelled ($l_1$, $l_4$, etc); (ii) scopal arguments are indicated by
holes (e.g., $h_2$ and $h_3$ in the quantifier {\em every}); (iii)
constraints on scope are expressed by the qeq conditions (roughly, $h
=_q l$ means that only zero or more quantifiers lie between the scopal
argument $h$ and the predication labelled by $l$ in any fully-resolved
logical form); and (iv) these scope
constraints allow for more than one way of `plugging' the holes with
labels.\footnote{The event variable in the predicate symbol
  corresponding to the adjective is motivated by copulas---{\em The
    cat is fat} has logical form 
  $\mbox{\_the\_q\_1}(x,\mbox{\_cat\_n\_1}(x),\mbox{\_fat\_j\_1}(e,x))$.}
\begin{examples}
\item   \label{eg:cat}
\begin{subexamples}
\item
Every fat cat chased some dog.
% \item
% $\mbox{\_every\_q\_1}(x, \mbox{\_fat\_j\_1}(e',x) \wedge
% \mbox{\_cat\_n\_1}(x),$\\
% \hspace*{0.5in}$\mbox{\_some\_q\_1}(y, \mbox{\_dog\_n\_1}(y),
% \mbox{\_chase\_v\_1}(e_{spast},x) )$\\
% $\mbox{\_some\_q\_1}(y, \mbox{\_dog\_n\_1}(y),$\\
% \hspace*{0.5in}$\mbox{\_every\_q\_1}(x, \mbox{\_fat\_j\_1}(e',x)
% \wedge 
% \mbox{\_cat\_n\_1}(x), 
% \mbox{\_chase\_v\_1}(e_{spast},x)))$
\item   $l_1\handel\mbox{\_every\_q\_1}(x,h_2,h_3)$\\
$l_4\handel\mbox{\_fat\_j\_1}(e',x)$\\
$l_4\handel\mbox{\_cat\_n\_1}(x)$\\
$l_5\handel\mbox{\_chased\_n\_1}(e,x,y)$\\
$l_6\handel\mbox{\_some\_n\_1}(y,h_7,h_8)$\\
$l_9\handel\mbox{\_dog\_n\_1}(y)$\\
$h_2 =_q l_4, h_7 =_q l_9$
\end{subexamples}
\end{examples}
(\ref{eg:cat}b) also exhibits {\sc mrs}' compact description of
conjunction through label sharing (see $l_4$ above), and the {\sc
  erg}'s naming convention for predicate symbols: each predicate is
constructed from the word lemma, its syntactic category and a sense
number, and the leading underscore identifies lexical predicates.

Conceptually, this {\sc mrs} is a partial description of all its
possible (fully specific) interpretations in context.  
These are expressed in a so-called {\em base language}, where each
predicate symbol in the {\sc mrs} is a proxy for one or more
base-language constructors of the same arity.
Assuming a very close
correspondence between the {\sc mrs} predicate symbols and the
base-language constructors they 
stand for, the {\sc mrs}
(\ref{eg:cat}b) (partially) describes both the base-language formulae in
Figure~\ref{fig:1}, shown as trees to emphasise that an {\sc mrs} is a
partial description of the formulae's {\em form}; not its
(base-language) entailments. 
In essence, the labels in the {\sc mrs} reify the
nodes in the tree, and each of these trees encodes one way of equating
holes with labels that respects qeq constraints and binding conditions
on the variables.

@@

The {\sc mrs} (\ref{eg:cat}c) is a notational variant of the {\sc
  rmrs} (\ref{eg:cat}d). {\sc rmrs}s offer a more factorised
representation: its predicate symbols are all unary and the other
arguments to the (base-language) constructor that it corresponds to
are represented by separate binary relations on the unique {\em
  anchor} of the predicate symbol ($a_1, a_{41},\ldots$) together with
variable and label equalities (e.g., $x=x_1$, $l_{41}=l_{42}$).
\begin{examples}
\item   [\ref{eg:cat}]
\begin{subexamples}
\item   [c]
$l_1:a_1\handel\mbox{\_every\_q\_1}(x_1),$\\
\hspace*{0.1in}$\mbox{RSTR}(a_1,h_2),
\mbox{BODY}(a_1,h_3)$\\ 
$l_{41}:a_{41}\handel\mbox{\_fat\_j\_1}(e'), \mbox{ARG1}(a_{41},x_2)$\\
$l_{42}:a_{42}\handel\mbox{\_cat\_n\_1}(x_3)$\\
$l_5:a_5\handel\mbox{\_chase\_v\_1}(e_{spast})$,\\
\hspace*{0.1in}$\mbox{ARG1}(a_5,x_4),
\mbox{ARG2}(a_5,x_5)$\\ 
$l_6:a_6\handel\mbox{\_some\_q\_1}(x_6)$,\\
\hspace*{0.1in}$\mbox{RSTR}(a_6,h_7),
\mbox{BODY}(a_6,h_8)$\\ 
$l_9:a_9\handel\mbox{\_dog\_n\_1}(x_7)$\\
$h_2=_q l_{42}, l_{41}=l_{42}, h_7 =_q l_9$\\
$x_1=x_2, x_2=x_3, x_3=x_4,$\\
$x_5=x_6, x_5=x_7$
\end{subexamples}
\end{examples}

This factored representation allows one to build semantic components
to shallow parsers where lexical or syntactic information is absent.
An extreme example would be a {\sc 
  pos} tagger: one can build its semantic component simply by deriving
lexical predicate symbols from the word lemmas and their tags and ensuring
that the predication for each word receives a unique label, anchor and
argument, as given in (\ref{eg:cat-pos}):
\begin{examples}
\item   \label{eg:cat-pos}
\begin{subexamples}
\item   Every\_{\sc at1} fat\_{\sc jj} cat\_{\sc nn1} chased\_{\sc
    vvd} some\_{\sc dd} dog\_{\sc nn1}
\item   $l_1:a_1:\mbox{\_every\_q}(x)$, \\
$l_{41}:a_{41}:\mbox{\_fat\_a}(e')$,\\
$l_{42}:a_{42}:\mbox{\_cat\_n}(x_3)$\\
$l_5:a_5:\mbox{\_chased\_v}(e_{past})$, \\
$l_6:a_6:\mbox{\_some\_q}(x_4)$, \\
$l_9:a_9:\mbox{\_dog\_n}(x_5)$
\end{subexamples}
\end{examples}
Semantic relations, sense tags and the arity of the base-language
constructors are
missing from (\ref{eg:cat-pos}b) because the {\sc pos} tagger doesn't
reveal information about syntactic constituency, word sense or lexical
subcategorisation respectively.  
But the {\sc rmrs}s (\ref{eg:cat}c) and
(\ref{eg:cat-pos}b) are entirely compatible, the former being more
specific than the latter.  At least, this holds so long as the
base-language constructors that the {\sc rmrs} predicate symbol
{\_name\_t\_s} corresponds to is a 
subset of those corresponding to the predicate symbol {\em \_name\_t},
for any word lemma {\em name}, {\sc pos} tag $t$ and sense $s$ (written 
$\mbox{{\em \_name\_t\_s}} \sqsubset \mbox{{\em name\_t}}$).

The values of non-scopal arguments are underspecified through the
absence of ARG relations and/or through the absence of
equalities and inequalities among the variables $x_1, x_2,\ldots$.
The need to
underspecify such semantic dependencies is required when
the shallow parser doesn't record syntactic dependencies.
Indeed, there is no single {\sc mrs} which is satisfied by both the
trees in Figure~\ref{fig:2}, assuming of course that disjunction is not part of
the {\sc mrs} language.  In words, these trees collectively specify
the (partial) semantic information that $x$ is either the second
argument (i.e., the direct object) or the third argument (i.e., the
indirect object) to {\em give}, but we don't know which.  
A shallow
processor that misses long distance dependencies in
(\ref{eg:which-dog}) cannot
discern that the bone is the second argument to {\em
  give} in (\ref{eg:which-dog}a), while the dog is the third argument
in (\ref{eg:which-dog}b).
\begin{examples}
\item   \label{eg:which-dog}
\begin{subexamples}
\item   To which dog did Kim give the bone?
\item   Which bone did Kim give the dog?
\end{subexamples}
\end{examples}
Thus to maximise the semantic contribution
of the parser while not overdetermining it, the phrases
{\em give the bone} vs.\
{\em give the dog} should whether the semantic index of the noun
phrase is the second or third argument to {\em give},
as shown in Figure~\ref{fig:2}.   This will be expressed in
{\sc rmrs} via the binary relation  $\mbox{{\em ARG}}_{\{2,3\}}(a,x)$
between the anchor $a$ of the predicate symbol for {\em give} and the
individual $x$.   More generally, $\mbox{{\em ARG}}_n(a,x)$ means that
$x$ is an argument to the predicate anchored at $a$, but its argument
position is unspecified.
